### 2019-09-19 boek.tex
update
updaze

@@

%% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH5: SCORPIUS       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
\label{chap:scorpius}
\begin{quote}
	\textbf{Abstract:} \blindtext
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, \ldots, De Preter, K., and Saeys, Y. SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development. \textit{Journal} vol, issue (2019), page--page. \doi{10.1101/079509v2}.
\newpage
\include{ch5_scorpius/content}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%      CH8: DISCUSSION      %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{General discussion} 
\chaptermark{General discussion.} % short title
\label{chap:discussion}
\include{ch8_discussion/content}

>>>>

% boxes around sections
\usepackage{tcolorbox}

@@

%% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}
\usepackage[open,openlevel=1]{bookmark}

@@

% si units and chemical stuff
\usepackage{siunitx}
\usepackage[version=3]{mhchem}

% command for adding comments
\newcommand{\mycomment}[1]{\noindent\fbox{\parbox{\textwidth}{\color{red}#1}}}

@@

% TIP FROM WOUTER: Per chapter, write 'update' reflecting on what came out of the paper

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH5: SCORPIUS       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{SCORPIUS: Fast, accurate, and robust single-cell pseudotime} 
\chaptermark{SCORPIUS: Fast, accurate, and robust single-cell pseudotime.} % short title
\label{chap:scorpius}
\begin{quote}
	\textbf{Abstract:} \blindtext
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, Saelens, W., Sichien, D., Tavernier, S., Janssens, S. Guilliams, M., Lambrecht, B., De Preter, K., and Saeys, Y. SCORPIUS improves trajectory inference and identifies novel modules in dendritic cell development. \textit{Journal} vol, issue (2019), page--page. \doi{10.1101/079509v2}.
\newpage
\graphicspath{{./ch5_scorpius/}}
\include{ch5_scorpius/content}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH8: Guidelines       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Essential guidelines for computational method benchmarking} 
\chaptermark{Essential guidelines for computational method benchmarking.} % short title
\label{chap:guidelines}
\graphicspath{{./ch8_guidelines/}}
\begin{quote}
	\textbf{Abstract:} In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses. Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis. However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results. Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.
\end{quote}
\vfill
Adapted from:\\
Weber, L. M., Saelens W., \textbf{Cannoodt, R.}, Soneson, C., Hapfelmeier, A., Gardner, P. P.,
Boulesteix, A., Saeys, Y., and Robinson, M. D.. 
Essential guidelines for computational method benchmarking. \textit{Genome Biology} 20, 125 (2019). \doi{10.1186/s13059-019-1738-8}.
\newpage
\include{ch8_guidelines/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%      CH8: DISCUSSION      %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Discussion and Future Perspectives} 
\chaptermark{Discussion and Future Perspectives.} % short title
\label{chap:discussion}
\include{ch8_discussion/content}

### 2019-09-19 ch1_introduction/content.tex
update

@@

\comment{
HT: I feel like you're only going to focus on preprocessing, when reading this section. Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way? \\
RC: Literally but succinctly state what the exact contributions of the thesis are, this is still very vague.
}

@@

\comment{
RC: Use crispr-cas to assert the impact of the discovery of DNA, not the FSVM spiral 
}

@@

\comment{use this reference? \cite{cornell_prebioticaminoacids_2019}}

@@

\subsection{The Central Dogma}
\comment{HT: feels like an enumeration, should make it more into a story}

@@

\comment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task. In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.

\comment{are dynamic processes defined? a dynamic process is just a biochemical reaction}

@@

\comment{TODO: download database and calculate some statistics?}

@@

\comment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
\subsection{Profiling single cells}
\comment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

@@

\comment{Rephrase the microscope sentence}

@@

On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
\comment{Demonstrate the masking effect of bulk analyses.}

@@

\comment{Make a better connection to the subsections below}

\subsection{Normalisation}
\comment{TODO}

@@

\comment{This section is not very interestingly written}

@@

@@

\comment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}

@@

\comment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}

@@

%Before embarking on this perilous adventure, I attempt to write a "Hippocratic oath for bioinformaticians", to guide me through this dissertation and not stray from the metaphorical path of righteousness.
%
%\subsection{Hippocratic oath for bioinformaticians}
%We proceed with caution in developing new tools. Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible. We value creating open-source software that works reliably but fails gracefully when it does not. We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects. % TODO: improve

@@

\comment{
	Submitted:\\
	RNA atlas,\\
	TRADE-seq: Trajectory-based differential expression analysis,\\
	SCENIC protocol
}

@@

>>>>

% INTRODUCTION CONTENT

@@

\mycomment{
HT: I feel like you're only going to focus on preprocessing, when reading this section. Maybe add a second paragraph mentioning that separating technical and biological noise from actual biological processes in the data is not the only challenge, and that there is also a necessity of new computational tools to start generating more complex models, integrating the information from single cell high sequencing experiments in a much more refined way? \\
RC: Literally but succinctly state what the exact contributions of the thesis are, this is still very vague.
}

@@

\mycomment{
RC: Use crispr-cas to assert the impact of the discovery of DNA, not the FSVM spiral 
}

@@

\mycomment{use this reference? \cite{cornell_prebioticaminoacids_2019}}

@@

\subsection{The Central Dogma}
\mycomment{HT: feels like an enumeration, should make it more into a story}

@@

\mycomment{do not only discuss differentiation, but instead any kind of developmental process; and mention 'developmental trajectory'.}
Cell differentiation is not an instant process; it is a continuous process in which a cell gradually produces the biochemical machinery required in order to fulfil a particular task. In this regard, it makes sense not only to reason about cell types but also about the transition states between cell types and the dynamic processes involved therein.

\mycomment{are dynamic processes defined? a dynamic process is just a biochemical reaction}

@@

\mycomment{TODO: download database and calculate some statistics?}

@@

\mycomment{TODO: explain regulation mechanisms such as transcription factor binding sites?}
\subsection{Profiling single cells}
\mycomment{This sections goes greatly into detail about IHC and cytometry. Can the merging of the 'single-cell' and 'omics' be explained without going so much into detail?}
In order to understand a biological process, it is often quite helpful to be able to profile (i.e. observe) the biomolecular components involved therein. The single-cell "omics" technologies which we have at our fingertips today originated from the convergence of two different fields, \emph{"single-cell"} and \emph{"omics"}.

@@

\mycomment{Rephrase the microscope sentence}

@@

On the other side of the spectrum are the so-called "omics" technologies. "Omics"\footnote{The etymology of "omics" is quite interesting \cite{yadav_wholenesssuffixomics_2007}.} is a collective term for profiling all molecules of a particular type in a high-throughput manner. There are many types of "omics", but the most commonly used are the following. In genomics, all of an organism's genes are studied -- its whole genome. Transcriptomics and proteomics study the organisms RNA transcripts and proteins, respectively. A notable downside of traditional omics technologies is that in order to capture enough material an ensemble of cells needs to be profiled, and thus only the average expression levels are returned; thereby granting the technology the name "bulk" omics. If a subset of these cells contains unique patterns in expression levels, this pattern will be masked in the bulk population and is thus undetectable. Specific examples of omics technologies are next-generation sequencing, which can be used to determine the DNA sequence of an organism, and RNA sequencing, which profiles the sequences of RNA transcripts. By mapping the sequences of RNA transcripts to genes in the organisms DNA, a gene expression profile can be obtained.
\mycomment{Demonstrate the masking effect of bulk analyses.}

@@

\mycomment{Make a better connection to the subsections below}

\subsection{Normalisation}
\mycomment{TODO}

@@

\mycomment{This section is not very interestingly written}

@@

%Technological advancements in single-cell omics allow studying a dynamic process in a high-throughput manner. 
%This raises concerns regarding biological fundamentals, such as how to define cell types or transitions between them \cite{etzrodt_quantitativesinglecellapproaches_2014,tanay_scalingsinglecellgenomics_2017}. 
%Trajectory inference (TI) methods aim to give insight into a dynamic process by inferring a trajectory from omics profiles of cells in which the dynamic process takes place \cite{cannoodt_computationalmethodstrajectory_2016}. 
%TI has two objectives: to reconstruct the topology of the dynamic process (e.g. is it linear, cyclical, bifurcating), and to determine the position of each cell along the topology. Some TI methods assumes that the user knows the topology beforehand and only focuses on ordering the cells along a predefined topology.

@@

\mycomment{Could still expand this section with pieces from the EJI paper, though it needs to be adapted strongly.}

@@

\mycomment{Get a better focus for the overall contributions of the thesis. For example, focus on datasets that contain developing cells, and that the tools we develop make this assumption.}

@@

@@

\mycomment{
	Submitted:\\
	RNA atlas,\\
	TRADE-seq: Trajectory-based differential expression analysis,\\
	SCENIC protocol
}

@@

		\cranpkg{Rdimtools} & ctb & CRAN & 7367 & Dimension Reduction and Estimation Methods \\

### 2019-09-19 ch8_discussion/content.tex
update

\blindtext

\section{Overview and conclusions of the presented work}

% notice that normalisation was absent from this dissertation.
%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.

\section{Future research directions}

>>>>

\section{Essential guidelines for computational method benchmarking}

%Before embarking on this perilous adventure, I attempt to write a "Hippocratic oath for bioinformaticians", to guide me through this dissertation and not stray from the path of righteousness.
%
%\subsection{Hippocratic oath for bioinformaticians}
%We proceed with caution in developing new tools. Its results should not only be convincing and easy to interpret, but also accurate, robust, and reproducible. We value creating open-source software that works reliably but fails gracefully when it does not. We acknowledge that writing automated tests is dull but necessary for maintaining long-term software projects. % TODO: improve

%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.

\section{dyngen discussion}

\section{dynbenchmark discussion}
\begin{itemize}
	\item Already outdated when the manuscript was published online
	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
	\item Include RNA velocity as inputs
\end{itemize}

\section{SCORPIUS discussion}
\begin{itemize}
	\item Extension to inferring non-linear trajectories, i.e. with principal graphs
\end{itemize}

\section{bred discussion}

\section{incgraph discussion}

\section{dyno discussion}
% notice that normalisation was absent from this dissertation.
